---
title: "Two sample MR"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# Two-Sample MR

Updated 22.05.2024

```{r}
working_dir <- "***"


exp_filename <- "***"
out_filename <- "***"

exp_name <- "XXX"
out_name <- "YYY"


p_lim <- 0.05

r2_lim <- 0.2
```

```{r}
library(TwoSampleMR)
library(data.table)
setwd(working_dir)

```

## Data preparation

Loading the exposure data 

```{r}

setwd(working_dir)
exp_raw_full <- fread(file =exp_filename)

```


Formatting exposure data and selecting significant SNPs (p-value)

```{r}
exp_raw <- subset(exp_raw_full,exp_raw_full$pval<p_lim)

exp_raw <- as.data.frame(exp_raw)
exp_raw$phen <- rep(exp_name, nrow(exp_raw))

exp_dat <- format_data( exp_raw,
    type = "exposure",
    snp_col = "SNP",
    beta_col = "beta",
    se_col = "se",
    effect_allele_col = "alt",
    other_allele_col = "ref",
    pval_col = "pval",
    eaf_col = "freq",
    phenotype_col = "phen",
    samplesize_col = "n"
)
```


## Clumping the exposure variables

```{r}
library(ieugwasr)
clumped_exp <- clump_data(exp_dat,clump_r2=r2_lim,pop="EAS")
```


## Outcome-data

```{r}

setwd(working_dir)
out_raw_full <- fread(out_filename)

out_raw <- subset(out_raw_full,out_raw_full$pval<p_lim)
out_raw <- as.data.frame(out_raw)
out_raw$phen <- rep(out_name, nrow(out_raw))

out_dat <- format_data( out_raw,
    type = "outcome",
    snp_col = "SNP",
    beta_col = "beta",
    se_col = "se",
    effect_allele_col = "alt",
    other_allele_col = "ref",
    pval_col = "pval",
    eaf_col = "freq",
    phenotype_col = "phen",
    samplesize_col = "n"
    
)
```

## Harmonizing data

```{r}
harmonized_data <- harmonise_data(clumped_exp,out_dat,action=1)
harmonized_data
```

## MR analysis

```{r}
scatter_name <- paste("scatter_plot_", exp_name, "_", out_name, ".pdf")

pdf(scatter_name)

res <- mr(harmonized_data, )

mr_scatter_plot(res[3, ], harmonized_data)
dev.off()
```

## Sensitivity analysis

### Heterogenity

```{r}
mr_pleiotropy_res <- mr_pleiotropy_test(harmonized_data)
mr_pleiotropy_res
```

## Single SNP and leave-one-out (not performed due to the large amount of SNPs)

### Single SNP MR

```{r}

res_single <- mr_singlesnp(harmonized_data)
res_single

```

### leave-one-out MR

```{r}

res_loo <- mr_leaveoneout(harmonized_data)
res_loo

```

## Visualization

### Scatter plot

```{r}
res <- mr(harmonized_data)
p1 <- mr_scatter_plot(res, harmonized_data)
p1[[1]]
```

### LOO visual

```{r}
res_loo <- mr_leaveoneout(harmonized_data)
p3 <- mr_leaveoneout_plot(res_loo)
p3[[1]]
```

### Funnel plot (single)

```{r}
res_single <- mr_singlesnp(harmonized_data)
p4 <- mr_funnel_plot(res_single)
p4[[1]]
```

## MR Steiger directionality test

```{r}
harmonized_data$"r.outcome" <- get_r_from_lor(
  harmonized_data$"beta.outcome",
  harmonized_data$"eaf.outcome",
  45383,
  132032,
  0.26,
  model = "logit",
  correction = FALSE
)

out <- directionality_test(harmonized_data)
out
```

## Identifying the lead SNPs

```{r}
sc_plot <- mr_scatter_plot(res, harmonized_data)
sc_plot[[1]]

# Smallest outcome pval
lead_out_pval <- which(harmonized_data$pval.outcome == min(harmonized_data$pval.outcome)  )
print("Lead outcome Pval:")
harmonized_data$SNP[[lead_out_pval]]
harmonized_data$pval.outcome[[lead_out_pval]]
harmonized_data[lead_out_pval, ]

# Smallest exposure pval
lead_exp_pval <- which(harmonized_data$pval.exposure == min(harmonized_data$pval.exposure)  )
print("Lead exposure Pval:")
harmonized_data$SNP[[lead_exp_pval]]
harmonized_data$pval.exposure[[lead_exp_pval]]
harmonized_data[lead_exp_pval, ]


# Smallest pval for single SNP analysis
print("Lead single SNP Pval:")
n_max <- nrow(res_single) -2
lead_single_ind <- which(res_single$p == min(res_single$p[1:n_max])  )
res_single$SNP[[lead_single_ind]]
res_single$p[[lead_single_ind]]
print(res_single[lead_single_ind, ])
```
